Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-05-13 content.tex
Add ERQ 6

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: Example Research Question 1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Now we can formulate the first example research question that \verb;pirouette; 
can answer.

\subsection{Question}

"What is the inference error made on phylogenies
created by a standard diversification model?"

\subsection{Answer}

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_1/true_tree.png}
  \caption{The Yule tree, as created by listing~\ref{lst:create_yule_tree}.}
  \label{fig:yule_tree}
\end{figure}

@@

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment_explicit},
  caption = {
    Create the experiments to answer research question 1, 
    using explicit arguments.
  }
]
experiment <- create_experiment(
  inference_conditions = create_inference_conditions(
    model_type = "generative", 
    run_if = "always"
  ), 
  inference_model = create_inference_model(
    tree_prior = create_yule_tree_prior(),
    clock_model = create_strict_clock_model(), 
    site_model = create_jc69_site_model()
  )
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment},
  caption = {
    Create the experiments to answer research question 1.
  }
]
experiments <- list(experiment)
\end{lstlisting}

@@

We will not use twinning here yet.  
All these arguments are bundled
and checked by \verb;create_pir_params;:

@@

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it from ERQ 3.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: Example Research Question 2}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$, by our definition, is non-standard, and hence 
usually it is not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is what standard tree prior should be 
used in the inference.
There are two ways to deal with this problem: hand-picking a tree
prior, or picking the best of a set of tree priors. 
Hand-picking a tree prior can be rational if one standard tree prior 
is more suitable than others for the inference, for example, when
the novel tree prior is most similar to a specific standard tree prior.
Instead of hand-picking, the best inference model from a list of candidate 
models can also be used in inference. Selecting the 'best' model in
\verb;pirouette; means selecting the model with the highest evidence (also known as marginal likelihood).

We show the procedure using \verb;pirouette; to answer the second research question.

\subsection{Question}

"What is the inference error made on a novel phylogeny when
using the best inference model?"

\subsection{Answer}

This time we start with a tree generated from an unknown 
diversification model, that has six taxa and a crown age of ten:

\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_2/tree_unknown.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing~\ref{lst:unknown_phylogeny}.
  }
\end{figure}

The first step is to simulate a DNA alignment from the 
given phylogeny. 
We will re-use the alignment parameters of the previous example 
as shown in listing~\ref{lst:create_alignment_params}.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model\_type} &
    \textbf{run\_if} & \textbf{do\_measure\_evidence} & 
    \textbf{inference model} \\ 
    \hline
    candidate & best candidate & TRUE & JC, strict, Yule \\
    candidate & best candidate & TRUE & JC, strict, BD   \\
    ...       & ...            & ... & ...              \\
    candidate & best candidate & TRUE & GTR, RLN, CCP    \\
    candidate & best candidate & TRUE & GTR, RLN, CEP    \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and inference model for ERQ 2.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR site model.
    RLN: relaxed log-normal clock model.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:RQ2}
\end{table}

@@

\begin{lstlisting}[
  language = R, 
  floatplacement = ht, 
  frame = single, 
  label = {lst:create_all_experiments},
  caption = Create all 40 candidate experiments.
]
experiments <- create_all_experiments()
\end{lstlisting}

For all the other steps we use the default setups. 
We can create the complete
\verb;pirouette; parameter set like this (which is the
same as listing~\ref{lst:create_pir_params}, but using the defaults):

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
as shown in Figure~\ref{fig:example_2}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_2/errors.png}
  \caption{
    ERQ 2: the inference error 
    when selecting the best inference model.
  }
  \label{fig:example_2}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: Example Research Question 3}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In ERQ 1 we measured the baseline error of the \verb;pirouette; method,
by using the same standard tree prior in both the generation of 
the original tree, as well as in the inference.
This baseline error distribution serves as a control,
to quantify the impact of a new (non-standard) tree prior.
To measure that impact, we compare the inference error 
from the new tree prior to the baseline error,
by using a twin tree (see subsection~\ref{subsec:twinning}).
In this ERQ, we hand-pick a standard tree prior $\mathit{p_{G}}$.

\subsection{Question}

"What is the inference error made from a phylogeny, 
when hand-picking a tree prior, compared to the background noise?"

\subsection{Answer}

Most of the settings for this ERQ are
the same as in ERQ 2: we will re-use the phylogeny from an 
unknown speciation model (listing~\ref{lst:unknown_phylogeny}), 
the alignment parameters (listing~\ref{lst:create_alignment_params}).

We create an experiment as in 
listing~\ref{lst:create_gen_experiment},
but with all defaults:

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment_defaults},
  caption = Create a list of one default generative experiment.
]
experiment <- create_gen_experiment()
\end{lstlisting}

Creating the complete set of experiments:

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_experiments},
  caption = Create a list of one experiment.
]
experiments <- list(experiment)
\end{lstlisting}

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
The output is shown in Figure~\ref{fig:example_3}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_3/errors.png}
  \caption{
    ERQ 3: the inference error made 
    when hand-picking a generative tree prior compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error,
    as measured on the twin tree.
  }
  \label{fig:example_3}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: Example Research Question 4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Rather than hand-picking a tree prior, 
one can select the best model from a set,
as in ERQ 2.
The resulting error distribution contains the full error accumulated in the pipeline, To quantify to what extent this error is due to the mismatch between 
the generative model $\mathit{G}$ and the selected inference model 
$\mathit{I}$, we can use twinning (see section \ref{subsec:twinning}: 'Twinning') to provide the baseline error.

\subsection{Question}

"What is the inference error from a phylogeny 
derived from an unknown speciation model,
when selecting the best inference model, 
in relation to the baseline error?"

\subsection{Answer}

Here, as in ERQs 2 and 3, we start with a phylogeny 
generated by an unknown speciation model (listing~\ref{lst:unknown_phylogeny}). 
The first step is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment_params}.

We specify that we want all 40 inference models to compete 
by calling \verb;create_all_experiments;,
as we previously did in listing~\ref{lst:create_all_experiments}.
We use again the default error measurement setup,
as shown in listing~\ref{lst:create_error_measure_params}.
To get an idea of a baseline error made by BEAST2, 
we also make use of the twinning option by using
the same call to \verb;create_pir_params; as shown in 
listing~\ref{lst:create_pir_params_with_twinning}.
We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_4}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_4/errors.png}
  \caption{
    ERQ 4: the inference error made 
    when selecting the 'best' tree prior compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error,
    as measured on the twin tree.
  }
  \label{fig:example_4}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: Example Research Question 5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

If we are moderately sure about hand-picking an inference model,
but would also like to check how well the hand-picked model compares
to the others, we can do both. Of course, as we are interested in the
baseline error for both the hand-picked and best candidate model,
we need to enable twinning as well. This example is a combination
of the previous two ERQs.

\subsection{Question}

"What is the error made by BEAST2 from a phylogeny, 
when the generative model is hand-picked, 
compared to candidates with similar inference models, 
in relation to the background error?"

\subsection{Answer}

In this example we use a tree generated by the Yule process.
Instead of using the common Yule tree as in ERQ 1, we will use
a Yule tree of low likelihood, 
i.e. the one already used in ERQs 2-4, 
by using the code shown in listing~\ref{lst:unknown_phylogeny}.
The first step is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment_params}.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model\_type} & 
    \textbf{run\_if} & 
    \textbf{do\_measure\_evidence} & 
    \textbf{inference model} \\ 
    \hline
    generative & always         & TRUE & JC, strict, Yule \\
    candidate  & best candidate & TRUE & JC, strict, BD \\
    candidate  & best candidate & TRUE & JC, strict, CCP \\
    candidate  & best candidate & TRUE & JC, strict, CEP \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the fourth research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:experiment_5}
\end{table}

In the second step we state our experiments. 
In this example we must specify both the generative and candidate experiments. 
Here we specify the generative model:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
generative_experiment <- create_gen_experiment()
\end{lstlisting}

We can specify all the candidates simply by calling 
\verb;create_all_experiments; and excluding the generative inference model,
resulting in 39 candidate experiments:

\begin{lstlisting}[
    language = R,
    floatplacement = H,
    frame = single
]
candidate_experiments <- create_all_experiments(
  exclude_model = generative_experiment$inference_model
)
\end{lstlisting}

We need to combine these experiments into one set of experiments:

\begin{lstlisting}[
    language = R,
    floatplacement = H,
    frame = single
]
experiments <- c(list(generative_experiment), candidate_experiments)
\end{lstlisting}

We use again the default error measurement parameters
created in listing~\ref{lst:create_error_measure_params}.
We enable twinning, by calling \verb;create_pir_params;
as shown in listing~\ref{lst:create_pir_params_with_twinning}.

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_5}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    Example 5: the inference error made, 
    when the generative model is hand-picked, 
    compared to the best of 39 candidate models, 
    each in relationship to the background error measured from the twin tree.
  }
  \label{fig:example_5}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Generative model only}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Now we can formulate the first example research question that \verb;pirouette; 
can answer: "What is the inference error made on phylogenies
created by a standard diversification model?"

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_1/true_tree.png}
  \caption{The Yule tree, as created by Listing~\ref{lst:create_yule_tree}.}
  \label{fig:yule_tree}
\end{figure}

@@

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment_explicit},
  caption = {
    Create the experiments to answer research question 1, 
    using explicit arguments.
  }
]
gen_experiment <- create_experiment(
  inference_conditions = create_inference_conditions(
    model_type = "generative", 
    run_if = "always"
  ), 
  inference_model = create_inference_model(
    tree_prior = create_yule_tree_prior(),
    clock_model = create_strict_clock_model(), 
    site_model = create_jc69_site_model()
  )
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment},
  caption = {
    Create the experiments to answer research question 1.
  }
]
experiments <- list(gen_experiment)
\end{lstlisting}

@@

All these arguments are bundled
and checked by \verb;create_pir_params;:

@@

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it in the last part of this example.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing to other candidate models}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$, by our definition, is non-standard, 
and hence usually it is not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is what standard tree prior should be 
used in the inference.
There are three ways to deal with this problem: hand-picking a tree
prior (as we did above), or picking the best of a set of tree priors, 
or to do both. 
Hand-picking a tree prior can be rational if one standard tree prior 
is more suitable than others for the inference, for example, when
the novel tree prior is most similar to a specific standard tree prior.
Instead of hand-picking, the best inference model from a list of candidate 
models can also be used in inference. Selecting the 'best' model in
\verb;pirouette; means selecting the model with the highest 
evidence (also known as marginal likelihood).

We show the procedure using \verb;pirouette; to answer the second research 
question: "What is the inference error made on a novel phylogeny when
using the best inference model, in comparison to a hand-picked model?

We will use the same tree as generated in \ref{lst:create_yule_tree}.

The first step is to simulate a DNA alignment from the 
given phylogeny. 
We will re-use the alignment parameters of the previous example 
as shown in Listing~\ref{lst:create_alignment_params}.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model\_type} &
    \textbf{run\_if} & \textbf{do\_measure\_evidence} & 
    \textbf{inference model} \\ 
    \hline
    generative & always         & FALSE & JC, strict, Yule \\
    candidate  & best candidate & TRUE  & JC, strict, BD   \\
    candidate  & best candidate & TRUE  & JC, strict, CBS  \\
    ...        & ...            & ...   & ...              \\
    candidate  & best candidate & TRUE  & GTR, RLN, CCP    \\
    candidate  & best candidate & TRUE  & GTR, RLN, CEP    \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and inference model for ERQ 2.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR site model.
    RLN: relaxed log-normal clock model.
    CBS: coalescent Bayesian Skyline tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:RQ2}
\end{table}

@@

\begin{lstlisting}[
  language = R, 
  floatplacement = ht, 
  frame = single, 
  label = {lst:create_all_experiments},
  caption = {
    Create all 40 candidate experiments, except for the inference model of the generative model.
  }
]
cand_experiments <- create_all_experiments(
  exclude_model = gen_experiment$inference_model
)
\end{lstlisting}

\begin{lstlisting}[
  language = R, 
  floatplacement = ht, 
  frame = single, 
  label = {lst:create_all_experiments},
  caption = {
    Create a collection of experiments, with 1
    generative model, and 39 candidate models
  }
]
experiments <- list(
  list(gen_experiment),
  cand_experiments
)
\end{lstlisting}

For all the other steps we use the default setups. 
We can create the complete
\verb;pirouette; parameter set like this (which is the
same as Listing~\ref{lst:create_pir_params}, but using the defaults):

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}),
as shown in Figure~\ref{fig:example_6}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/errors.png}
  \caption{
    ERQ 6: the inference error
    for both the generative and best candidate inference models.
  }
  \label{fig:example_6}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing to background noise}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Up until now, we measured the baseline error of the \verb;pirouette; method,
by using the same standard tree prior in both the generation of 
the original tree, as well as in the inference.
This baseline error distribution serves as a control,
to quantify the impact of a new (non-standard) tree prior.
To measure that impact, we compare the inference error 
from the new tree prior to the baseline error,
by using a twin tree (see subsection~\ref{subsec:twinning}) for
both the generative and best candidate model.

The research question this example answers is:
"What is the inference error made from a phylogeny,
for both a generative model and a best candidate model, 
compared to the background noise?"

Most of the settings for this example are
the same as in ERQ 2: we will re-use the phylogeny as 
before (Listing~\ref{lst:create_yule_tree}), 
as well as the alignment parameters (Listing~\ref{lst:create_alignment_params}).
Also the experiments are the same as in 
Listing~\ref{lst:create_all_experiments}.

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}),
The output is shown in Figure~\ref{fig:example_5}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error
  }
  \label{fig:example_5}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interpretation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
